Polymethylation scores paper

### 2021-11-02 Papers/Polymethylation_scores/04_background.Rmd
responses to kelly edits november2

Maternal prenatal smoking is a public health concern: in 2018, 11% of pregnant US women aged 15-44 reported any past-month cigarette use [@SAMHSA2018]. Exposed children are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu]. Measuring prenatal maternal smoking behavior is challenging. Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua]. The gold-standard for smoking measurements, serum cotinine, has a short half-life of nine hours in pregnant women [@Dempsey:2002wr]. Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures. Developing a portable and reliable biomarker of prenatal maternal smoke exposure would have important implications for research and clinical practice.

Differences in DNA methylation signatures of cordblood and placental tissues by prenatal maternal smoke exposure exist in primarily European cohorts of newborns [@Joubert:2016ue; @Everson:2021td; @Cardenas2019]. Blood DNA methylation signatures in cohorts of older children and adults also differ by prenatal maternal smoking exposure [@Tehranifar:2018tb ; @Richmond:2018tu]. Yet few studies have examined associations at different ages in the same participants. Additionally, while associations are well-replicated in cohorts of European ancestry, only a few, small sample size studies are in non-European ancestry populations. One 2016 study included 186 participants of African or admixed ancestry [@Ladd-Acosta:2016aa], while a 2018 study of 89 participants had 60% of African or admized ancestry [@Tehranifar:2018tb]. A larger 2019 study of 572 Latino children replicated associations with prenatal maternal smoking at 7 out of 148 tested *a priori* CpG sites from Joubert et al. at a FDR-adjusted *P*-value\<0.05 [@Neophytou:2019ty; @Joubert:2016ue]. Thus, while DNA methylation is associated with prenatal maternal smoking, the consistency of this association across age and ancestry is understudied.

The association between prenatal maternal smoking and DNA methylation suggests DNA methylation could provide a reliable and specific biomarker of prenatal maternal smoke exposure. Importantly, changes in DNA methylation are specific to smoke exposure *in utero*. Previous work has found reasonable performance of cord-blood accuracy ranging from 81% to 91%) [@Reese:2017ts; @Ladd-Acosta:2016aa]. A 2018 study found that DNA methylation based scores from adult peripheral blood could even predict predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu]. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across age and ancestry has not been evaluated. Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance. Additionally, previous work has relied on blood and placental samples. DNA methylation varies across cell-types and tissues, and epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples[@Everson:2021td]. Cord blood or placenta samples are unlikely to be available in a clinical setting. While peripheral blood is a more realistic clinical sample, saliva is even easier to collect. To our knowledge, no study has evaluated salivary DNA methylation as a biomarker for prenatal maternal smoking.

DNA methylation changes hold promise as a biomarker for prenatal maternal smoking. However, before translation to clinical applications, we must understand how and if the signal varies across age, ancestry and tissue sample. The goal of this study was to assess the *in vitro* potential of salivary DNA methylation biomarkers for prenatal maternal smoking. We tested associations between prenatal maternal smoke exposure and saliva DNA methylation biomarkers from the Illumina 450K chip, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites using longitudinal samples from the Fragile Families and Child Wellbeing study. We performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a consistent biomarker of prenatal maternal smoking across populations and time.

>>>>

Maternal prenatal smoking is a public health concern. In 2018, 11% of pregnant women in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018]. Children exposed to cigarette smoking in utero are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu]. Measuring prenatal maternal smoking behavior is challenging. Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua]. The gold-standard for smoking measurements, serum cotinine levels, have a short half-life of nine hours in pregnant women [@Dempsey:2002wr]. Further, serum cotinine measures during pregnancy are rarely available outside of birth cohorts, limiting clinical application as a marker of prenatal exposures. Developing a portable and reliable biomarker of prenatal maternal smoke exposure would have important implications for research and clinical practice.

Maternal cigarette smoking during pregnancy is associated with reproducible DNA methylation signatures, measured in infant cord blood and placental tissues, and primarily tested in European cohorts [@Joubert:2016ue; @Everson:2021td; @Cardenas2019]. Prenatal maternal cigarette smoking exposure is associated with postnatal DNA methylation signatures in blood from cohorts of older children and adults[@Tehranifar:2018tb ; @Richmond:2018tu]. Few studies have considered the persistence of the association between maternal prenatal cigarette smoking and DNA methylation at different ages in the same individuals. Additionally, little work has considered the portability of the association between maternal prenatal cigarette smoking and DNA methylation across tissues and ancestry groups. DNA methylation drives cell differentiation and cell type proportions differ across tissue types. Epigenetic markers of prenatal maternal smoking are known to differ between cord blood and placental samples [@Everson:2021td]. Cord blood and placental samples are unlikely to be available in clinical settings. While peripheral blood is a more realistic clinical samples, saliva is even easier to collect. To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation. Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017], and DNA methylation signatures of own-smoking differ across ancestry groups [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012]. Only a few studies have examined associations between prenatal maternal cigarette smoking and DNA methylation in non-European ancestry populations. Among 954 infants, 70% of whom had mothers identifying as Black, prenatal maternal smoking was associated with cord blood DNA methylation at 38 CpG sites [@Xu2021]. The direction of these association between prenatal maternal cigarette smoking and DNA methylation at these sites were generally similar in children of Black and non-Black participants [@Xu2021]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking was associated with DNA methylation in peripheral blood samples but ancestry-stratified analyses were not performed [@Ladd-Acosta:2016aa]. Similarly, among 89 middle-aged women, of whom reported African American or Hispanic ethnicity, prenatal maternal cigarette smoking was associated with DNA methylation at 17 of 190 tested CpG sites [@Tehranifar:2018tb]. Adjusting for ancestry did not substantially affect the association between prenatal maternal smoking and DNA methylation at any of the sites, although the authors note that African American women had on average 2--5% higher mean DNA methylation on the absolute scale as compared with White and Hispanic women at 5 *CYP1A1* CpG sites [@Tehranifar:2018tb]. Of 148 CpG sites selected based on their association prenatal maternal smoking among primarily European children, 7 CpG sites were associated with prenatal maternal cigarette smoking in a cohort of 572 Latino children [@Neophytou:2019ty; @Joubert:2016ue]. Thus, while prenatal maternal cigarette smoking is associated with DNA methylation, the consistency of this association across age, tissue and ancestry is understudied.

An important next step after association testing is the evaluation of DNA methylation as a biomarker of prenatal maternal smoke exposure, which is often done by comparing DNA methylation based classification of maternal smoking behavior to self-reported behavior. <!--#DEFINE POLYMETHYLATION SCORES HERE?-->DNA methylation signatures from cord blood classified maternal smoking behavior with accuracy of 91% in a primarily European cohort [@Reese:2017ts]. Among 572 3-5 year old children, 186 of whom were of African or admixed ancestry, prenatal maternal smoking behavior could also be classified with an accuracy of 81% when using a support vector machine approach and DNA methylation at 26 sites measured in peripheral blood at a single time point [@Ladd-Acosta:2016aa]. Among middle-aged adults, peripheral blood from a single time point could even predict prenatal smoke exposure (30 years previously) with an area under the curve of 0.72 (95% confidence interval 0.69, 0.76) [@Richmond:2018tu]. However, the consistency of DNA methylation as a biomarker for prenatal maternal smoking across tissue, age and ancestry has not been evaluated. Few studies have compared the effectiveness of DNA methylation as biomarker of prenatal maternal smoking in the same individuals at different times or evaluated if the inclusion of multiple time points improves performance. No study has evaluated the accuracy of salivary DNA methylation as a biomarker for prenatal maternal smoking.

DNA methylation changes hold promise as a biomarker for prenatal maternal smoking. However, before translation to clinical applications, we must understand how and if the signal varies across age, ancestry and tissue sample. In the Fragile Families and Child Wellbeing study, a diverse longitudinal cohort of children, we aimed to assess the potential of salivary DNA methylation biomarkers for prenatal maternal cigarette smoking. We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including epigenetic clocks, polymethylation scores, and individual *a priori* CpG sites. We then performed age- and ancestry-stratified analyses to test the hypothesis that DNA methylation could serve as a persistent and portable biomarker of prenatal maternal smoking.

### 2021-11-02 Papers/Polymethylation_scores/06_discussion.Rmd
responses to kelly edits november2

In the longitudinal Fragile Families and Child Wellbeing birth cohort, we found an association between prenatal maternal smoking and DNA methylation in sequential saliva samples from ages 9 and 15. Polymethylation scores were associated with prenatal maternal smoking across ages and ancestry while global methylation and epigenetic clocks were not, demonstrating a specific and persistent signal of prenatal smoke exposure in DNA methylation. Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking. This accuracy was more consistent across time than when using some individual *a priori* CpG sites.

Our findings fit within the previous literature on associations between prenatal maternal smoking and DNA methylation. We replicate top hits from previous epigenome wide association studies, including a large metaanalysis of DNA methylation in blood samples [@Joubert:2016ue]. Although there are no previous studies of prenatal maternal smoking and saliva methylation, our findings fit with findings on the consistency of DNA methylation signatures of own-smoking between saliva and blood samples [@Dawes2019; @Philibert2020].

We contribute to the literature on DNA methylation and prenatal maternal smoking by evaluating the consistency of this association in saliva and across ages and ancestries. Polymethylation scores built using coefficients from metaanalysis of blood DNA methylation still associated with prenatal maternal smoking in our independent population of saliva samples. Similarly, even though the source population for construction of the polymethylation scores was primarily of European ancestry, an association with prenatal maternal smoking was apparent in our primarily African ancestry population. Effect estimates were consistent across ancestries at age 9, although Hispanic and European ancestry age 15 samples had a lower effect estimate. This could reflect higher unreported smoking initiation in European and Hispanic ancestry US teens than African ancestry teens. Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016]. Own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification. This could also explain the observed attenuation of hypomethylation at AHRR:CpG05575921 by age 15 in our sample, as methylation at this probe is known to vary by own-smoking exposure as well[@Dawes2019; @Philibert2020].

Our findings also fit with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure. Accuracy of the polymethylation scores in our sample (Age 9: 0.67 and Age 15: 0.66) was worse than the performance of biomarkers using cord blood (AUCs ranging from 0.82 to 0.97) [@Reese:2017ts; @Ladd-Acosta:2016aa]. Differences in methylation patterns across tissues and over time could be responsible for this diminished accuracy. However, our accuracy was comparable to accuracy when using peripheral blood from older adults (AUC 0.72) [@Richmond:2018tu]. The diminished accuracy from cord blood may be an acceptable trade-off considering saliva is a more readily available tissue for epidemiological and clinical applications.

Our analysis answers pertinent questions for the development of a clinically applicable methylation biomarker of prenatal maternal smoking. We show that DNA methylation scores calculated from children's saliva can accurately classify prenatal maternal smoke exposure in non-European ancestry children. Other bomarkers and risk scores are known to differ across ancestry groups, potentially contributing to and exacerbating health disparities. As new biomarkers are developed, understanding their consistency across ancestry groups is critical. We also demonstrated that accuracy is not substantially affected when using coefficients from newborn cordblood (no cell type control, age 9: 0.62, age 15: 0.61) vs older children peripheral blood (no cell type control age 9: 0.63, age 15: 0.60). However, using coefficients from models that incorporated cell-type control at the time of calculation did result in better accuracy, suggesting cell-type control should be instituted both when creating weights and when applying scores to new population. Including samples from both ages did not improve accuracy, thus repeated samplings may be unnecessary for clinical applications. Additionally, accuracy was more consistent when using polymethylation scores than using the single top-hit from Joubert et al, AHRR:CpG05575921 [@Joubert:2016ue]. Salivary AHRR:CpG05575921 methylation was less accurate in our sample at predicting prenatal maternal smoke exposure than when it has been used to classify own smoking in blood or saliva [@Philibert2020]. While using a single CpG site as a biomarker may be clinically pragamatic, single sites may be less robust to time-since-exposure and new environmental exposures. Alternatively, incorporating information from across several CpGs in a polymethylation score may be more robust to these .

Our analysis is not without its limitations. We used maternal self report of prenatal maternal smoker as serum cotinine levels were not available. Due to social desirability bias, this could result in exposure misclassification. We would expect this to bias our results towards the null. In any analysis of a prenatal exposure and postnatal outcome, there is the possibility of selection bias into the cohort due to live birth bias. Selection bias is also possible due to loss-to-follow-up between birth and age 9. Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility of residual confounding in this observational cohort. However, the longitudinal design with repeated measures, exposure-outcome temporality and large sample size are strengths of our analysis.

>>>>

In the longitudinal Fragile Families and Child Wellbeing birth cohort, we observed prenatal maternal smoking was associated with DNA methylation in children's saliva samples from ages 9 and 15. Prenatal maternal smoking was associated with polymethylation scores across strata of child age and ancestry. Global methylation and epigenetic clocks were not associated with maternal smoking exposure demonstrating a specific and persistent signal of prenatal smoke exposure in DNA methylation. Polymethylation scores had reasonable accuracy for classifying prenatal maternal smoking. This accuracy was more consistent across time than when using individual *a priori* CpG sites.

Our findings are consistent with the previous literature on associations between prenatal maternal smoking and DNA methylation. We replicated top hits from previous epigenome wide association studies, including a large meta-analysis of DNA methylation in newborn blood samples [@Joubert:2016ue]. There are no previous studies of prenatal maternal smoking and saliva DNA methylation. However, previous work has that the majority of CpG sites were similarly methylated in blood and saliva (reviewed in @Langie2017). DNA methylation at AHRR:cg05575291 can accurately classify own--smoking behavior in both blood (area under the curve 0.995) and saliva (area under the curve 0.971) [@Dawes2019; @Philibert2020].

We advance the literature on prenatal maternal smoking and DNA methylation by evaluating the persistence of the association as children age and its portability across tissue and ancestry. Polymethylation scores built using coefficients from meta-analysis of cord blood DNA methylation from primarily European-ancestry newborns were still associated with prenatal maternal smoking in our independent saliva samples from a diverse cohort at ages 9 and 15 [@Joubert:2016ue]. The portability of other risk scores, such as polygenic risk scores, across ancestries is a complex research area [@Martin2019; ] and evaluating the portability of epigenetic summary measures has been identified as a key area for evaluation [@Hüls2020]. In this case, the polymethylation score for prenatal maternal smoking appears to be portable across ancestry groups. Effect estimates were consistent across ancestries at age 9, although Hispanic and European genetic ancestry samples had lower effect estimates at age 15. This could reflect higher unreported smoking initiation in European and Hispanic ancestry teens in the United States than African ancestry teens. Genetic ancestry correlates with race, and White and Latino teens have much higher rates of teen smoking and earlier ages at initiation than Black children [@El-Toukhy2016]. Though we excluded children who reported own-smoking from our analytic sample, under-reporting of own-smoking could influence DNA methylation at ages 9 and 15, creating outcome misclassification. This could also explain the observed attenuation the effect on AHRR:cg05575921 methylation by age 15 in our sample, as DNA methylation at this probe is known to vary by personal cigarette smoking[@Dawes2019; @Philibert2020].

Our findings are also consistent with previous work on DNA methylation as a biomarker for prenatal maternal smoke exposure. Prenatal maternal smoking classification using saliva DNA methylation in our sample of 9 and 15 year olds () performed comparably to classification using peripheral blood from older adults () [@Richmond:2018tu]. However, previous biomarkers of prenatal maternal smoking using DNA methylation of cord blood from newborns () and peripheral blood from 3-5 year old children () more accurately classified prenatal maternal smoke exposure [@Reese:2017ts; @Ladd-Acosta:2016aa]. Differences in DNA methylation patterns across tissues and over time may influence the performance of DNA methylation biomarkers. The similarity in performance between saliva DNA methylation in our study and peripheral blood DNA methylation in older adults is encouraging for the use of saliva DNA methylation as a readily accessible clinical sample.

Our analysis contributes to the development of an accurate methylation biomarker for prenatal maternal smoking by evaluating the impact of specific methodological choices on accuracy of classification. DNA methylation biomarkers may be especially susceptible to confounding by subject age at methylation measurement, tissue of measurement, and cell-type proportion [@Hüls2020]. Accuracy of polymethylation score classification was similar when using coefficients from cord blood vs peripheral blood samples from older children. Accuracy was not improved by using two time-points of methylation measurement. Accuracy was improved when polymethylation scores used coefficients which incorporated cell-type control. Our analysis therefore suggests that cell-type control when both generating and applying coefficients for polymethylation scores may positively influence accuracy, although more work comparing coefficients from different populations is still needed.

Our analysis suggests that polymethylation scores may be more accurate than using single CpG sites as biomarkers. AHRR:cg05575921 is a consistent marker of prenatal maternal smoke exposure in meta-analyses of newborn cord blood[@Joubert:2016ue] and a very accurate DNA methylation marker of own smoking [@Philibert2020]. However, salivary AHRR:cg05575921 was not a persistent marker of prenatal maternal smoking in an analysis of middle-aged adult women [@Richmond:2018tu]. In our sample, salivary AHRR:CpG05575921 methylation was less accurate at predicting prenatal maternal smoke exposure than polymethylation scores. The accuracy of AHRR:CpG05575921 and other single CpG site biomarkers may be influenced by time-since-exposure and new environmental exposures. Incorporating information across multiple sites of DNA methylation may yield a biomarker more robust to these influences.

Our analysis is not without its limitations. We used maternal self report of prenatal maternal smoker as serum cotinine levels were not available. Due to social desirability bias, this could result in exposure misclassification. We would expect this to bias our results towards the null. In any analysis of a prenatal exposure and postnatal outcome, there is the possibility of selection bias into the cohort due to live birth bias. Selection bias is also possible due to loss-to-follow-up between birth and age 9. Additionally, while we controlled for postnatal secondhand smoke exposure and excluded children who reported any own-smoking, we cannot exclude the possibility of residual confounding in this observational cohort.

However, our analysis also has several strengths. We analyzed samples from a large sample of diverse participants underrepresented in genetic and epigenetic research [@Sirugo2019; @Bentley2017]. While our exposure measurement of prenatal maternal smoking was self reported, it was assessed prospectively and preceded outcome measurements. We analyzed repeated measures of DNA methylation with reproducible array measures conducted in a single batch. We tested associations between prenatal maternal smoking and multiple DNA summary measures to evaluate the specificity of the polymethylation scores. In sensitivity analyses, we adjusted for other prenatal exposures and postnatal smoke exposure to examine the specificity of the biomarker to nature and timing of exposure.

### 2021-11-02 Papers/Polymethylation_scores/07_conclusions.Rmd
responses to kelly edits november2

Our analysis contributes to the clinical implementation of a DNA methylation-based biomarker of prenatal maternal smoking. We showed that DNA methylation in saliva, a clinically practical tissue, had strong associations with and reasonable accuracy for prenatal maternal smoke exposure. Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across ancestry-groups, an important consideration for health equity. The development and application of biomarker for prenatal maternal smoke exposure has important implications for epidemiological research and clinical practice. Given the difficulty of measuring prenatal maternal smoke exposure, such a biomarker could allow for confounder control in research areas where such control is currently impossible. Given the prevalence and negative health consequences of prenatal maternal smoke exposure, such a biomarker could be used to identify exposed individuals who might especially benefit from certain interventions.

>>>>

In a large, prospective study of diverse participants, we showed that DNA methylation in children's saliva had strong associations with and reasonable accuracy for prenatal maternal smoke exposure. Further, we demonstrated that polymethylation scores could be applied as a biomarker of prenatal maternal smoke exposure across ancestry-groups, an important consideration for the equitable biomarker development. The development and application of biomarker for prenatal maternal smoke exposure has important implications for epidemiological research and clinical practice. Given the difficulty of measuring prenatal maternal smoke exposure, such a biomarker could allow for confounder control in research areas where such control is currently impossible. Prenatal maternal smoke exposure is prevalent and has negative health consequences, thus an exposure biomarker could be used to provide support and health interventions for children.

### 2021-11-02 Papers/Polymethylation_scores/08_methods.Rmd
1st rough draft
responses to kelly edits november2

---
output: html_document
bibliography: "../Methylation_SmokingBib.bib"  
---

# Methods

## Cohort

The Fragile Families and Child Wellbeing Study (FFCW) is a nationally-representative cohort of nearly 5,000 children born in 20 U.S. cities between 1998 and 2000 [@Reichman:2001wv]. FFCW sampled births using a three-stage stratified random sample design, described in detail elsewhere [@Reichman:2001wv]. FFCW excluded the following births from the study: those with parents who planned to place the child for adoption, those where the father was deceased, those who did not speak English or Spanish well enough to be interviewed, those where the mothers or babies were too ill to complete the interview and those where the baby died before the interview could take place. At some of the hospitals, parents who were under 18 were not eligible to be interviewed due to hospital prohibition. Children were followed from birth until age fifteen. At the time of birth, medical record information was extracted from the hospital at which the mother gave birth. Interviews were conducted with the mother at birth in person and with father in person or over the telephone if he was not present. Interviews were also conducted with the mother and father when the child was one, three, five, and nine via telephone. Additionally, at ages three, five and nine, a separate primary caregiver survey was administered to the individual with whom the child was living either over the telephone or in the home. The child was interviewed at age nine in the home and at age fifteen by telephone or in the home. At ages nine and fifteen a saliva sample was taken from the child for DNA and DNA methylation measurements [@Reichman:2001wv].

## Prenatal maternal smoking measurement

## Covariates

## DNA methylation measurement

## Sample filtering

## Probe filtering

## Methylation summary measures

We created several methylation summary measures. Global methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set. Methylation clocks were calculated for each sample according to the methods of --INSERT HERE--. To create polymethylation scores, 6,000 (CHECK) CpG sites with regression coefficients reported in @Joubert:2016ue were multiplied by methylation beta values from FFCW and summed. We calculated polymethylation scores using regression coefficients from 4 different regressions reported in @Joubert:2016ue (LIST) and 3 different methods of transforming the methylation beta values (no transformation, mean center, and z-score standardization).

## Statistical analysis

>>>>

---
output:
  word_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
---

## Methods

### Cohort

The Fragile Families and Child Wellbeing Study is a birth cohort of nearly 5,000 children born in 20 cities in the United States between 1998 and 2000 [@Reichman:2001wv]. Participants were selected at delivery using a three-stage stratified random sample design which oversampled unmarried mothers by a ratio of 3:1 [@Reichman:2001wv]. Participants were excluded on the following criteria: those with parents who planned to place the child for adoption, those where the father was deceased, those who did not speak English or Spanish well enough to be interviewed, births where the mothers or babies were too ill to complete the interview, and those where the baby died before the interview could take place. Children were followed longitudinally with assessments at ages 1, 3, 5, 9 and 15; additional follow up is ongoing. Assessments included medical record extraction, biosample collection, in-home assessments, and surveys of the mother, father, primary caregiver, teacher and child. At ages nine and fifteen a saliva sample was taken from the child [@Reichman:2001wv]. A subsample of the Fragile Families cohort was selected for saliva DNA and DNA methylation processing. <!--#This subsampling was .... PUT SUBSAMPLE technique here -->

### Covariates and exposure measurement

Demographic and prenatal maternal substance use variables were derived from maternal self-report questionnaire data at baseline (child's birth).

Maternal covariates included maternal income to poverty ratio at baseline, prenatal smoking, prenatal maternal drug and alcohol use, and postnatal maternal or primary caregiver smoking. Maternal income to poverty ratio is a constructed variable of the ratio of total household income (as self-reported by the mother) to the official poverty thresholds designated by the United States Census Bureau for the year preceding the interview. At baseline, mother's answered categorical questions about their prenatal smoking, drug and alcohol use. For maternal prenatal smoking, mother's were asked

> During your pregnancy, how many cigarettes did you smoke? Did you smoke...:
>
> 2 or more packs a day
>
> 1 or more but less than 2 packs per day
>
> Less than 1 pack a day
>
> None

Few participants reported smoking a pack or more a day, thus we dichotomized to any vs no prenatal maternal smoking. For maternal prenatal drug and alcohol use the mothers were asked

> During your pregnancy how often did you use drugs/drink alcohol (respectively):
>
> Never
>
> Less than 1 time per month
>
> Several times per month
>
> Several times per week
>
> Every day

When the child was 1, 5, 9 and 15 years of age, primary caregivers responded to questions about maternal and in-home smoking. To encapsulate general early childhood smoke exposure, we created a binary variable for postnatal exposure at ages 1 or 5. To encapsulate recent postnatal smoke exposure, we used a categorical variable for packs per day (no smoking, less than one pack/day, one or more packs/day) in the month prior to the age 9 and 15 interview.

Child covariates included child sex, child report of personal cigarette smoking, and child genetic ancestry. Mothers reported sex (male/female) of their child at baseline. At ages 9 and 15, the child responded to interview questions about ever smoking a cigarette.

Child genetic ancestry was calculated. Principal components of child genetic ancestry were calculated from genetic data (CHIP?) from child saliva samples.

### DNA methylation measurement

Salivary samples from the children were collected at ages nine and fifteen using the Oragene•DNA sample collection kit (DNA Genotek Inc., Ontario). Saliva DNA was extracted manually using DNA Genotek's purification protocol using prepIT L2P. DNA was bisulfite treated and cleaned using the EZ DNA Methylation kit (Zymo Research, California). Samples were randomized and plated across slides by demographic characteristics. Saliva DNA methylation was measured using the Illumina HumanMethylation 450k BeadArray [@Sandoval2011] and imaged using the Illumina iScan system. All samples were run in a single batch to minimize technical variability.

DNA methylation image data were processed in R statistical software (3.5) using the minfi package [@Aryee2014]. The red and green image pairs (n=1811) were read into R and the minfi *preprocessNoob* function was used to normalize dye bias and apply background correction. Further quality control was applied using the ewastools packages [@Heiss2018]. We dropped samples with \>10% of sites have detection p-value \>0.01 (n=43), sex discordance between DNA methylation predicted sex and recorded sex (n=20), and abnormal sex chromosome intensity (n=3). CpG sites were removed if they had detection p-value \>0.01 in 5% of samples (n=26,830) or were identified as cross-reactive (n=27,782) (Figure 1) [@Chen2013]. Relative proportions of immune and epithelial cell types were estimated from DNA methylation measures using a childhood saliva reference panel [@Middleton2021].

We created polymethylation scores for prenatal maternal smoke exposure. From an independent meta-analysis of prenatal smoke exposure and newborn DNA methylation, we extracted the regression coefficients of 6,074 CpG sites associated with prenatal maternal smoking at a false discovery ratio corrected $P$ value \<0.05 [@Joubert:2016ue]. We mean-centered the DNA methylation beta values in our study, weighted them by the independent refression coefficience and took the sum. We calculated polymethylation scores using regression coefficients from 4 different regressions (sustained smoking exposure and DNA methylation in newborn cordblood with cell-type control, sustained smoking exposure and DNA methylation in newborn cordblood without cell-type control, sustained smoking exposure and DNA methylation in peripheral blood from older children, any smoking exposure and DNA methylation in newborn cordblood) [@Joubert:2016ue]. Global DNA methylation was calculated for each sample as the mean methylation value of each sample across the cleaned probe set. Pediatric epigenetic age was calculated for each sample using the coefficients and methods provided by the creators (see [<https://github.com/kobor-lab/Public-Scripts>](https://github.com/kobor-lab/Public-Scripts){.uri}) [@mcewen2019]. The GRIM age clock, including the smoking pack-years subscale, was calculated as previously described [@lu2019].

Single *a priori* CpG sites were selected from the previous large meta-analysis of prenatal maternal smoking and DNA methylation in children's cord and peripheral blood samples [@Joubert:2016ue].

Surrogate variables were calculated from DNA methylation data using the function sva from the R package sva version 3.38.

### Statistical analyses

#### Inclusion exclusion criteria

From the 1692 samples from 884 unique individuals with quality controlled DNA methylation data, we further excluded any samples missing data on: maternal prenatal smoking (0 samples), alcohol (4 samples) or other drug use (6 samples), maternal income to poverty ratio (0 samples), maternal postnatal smoking data ( samples). We also excluded samples with missing child sex (0 samples), child age (0 samples) or genetic data ( samples). Finally, if a child reported ever smoking a cigarette at age 9, we excluded their samples. Age 15 samples from children who reported ever smoking a cigarette at age 15 were also excluded.

#### Primary model

In our primary model we adjusted for maternal income to poverty ratio at baseline, child sex, child age, plate from DNA methylation processing, and estimated immune cell proportion estimated from DNA methylation. In nonstratified models, we adjusted for the first two components of genetic ancestry from principal component analysis. In ancestry-stratified models, the first two principal components from principal component analysis run within each ancestry strata were used. While child sex, child age, plate from DNA methylation processing and immune cell proportions are not confounders (as they cannot casually effect prenatal maternal smoke exposure), these variables can strongly effect DNA methylation and so were adjusted for as precision variables.

#### Sensitivity analyses

In sensitivity analyses we performed additional adjustments: 1) models adjusting for prenatal drug and alcohol use, 2) models adjusting for prenatal drug and alcohol use and postnatal maternal/primary caregiver smoking, 3) models adjusting for surrogate variables calculated from DNA methylation data.

Code to perform all analyses is available (www.github.com/bakulskilab)
